---
title: "Analysing Nanobody CDR3 Sequences with abseqr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing Nanobody CDR3 Sequences with abseqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi = 150
)
```

## Introduction

Nanobodies are single-domain antibody fragments (VHH) derived from camelid
heavy-chain-only antibodies. Their complementarity-determining region 3
(CDR3) is the primary determinant of antigen binding specificity and tends to
be longer and more structurally diverse than the CDR3 of conventional
antibodies.

This vignette demonstrates how to use `abseqr` to:

1. Parse a FASTA file of nanobody sequences into a tidy data frame
2. Extract CDR3 regions using conserved flanking motifs
3. Summarise and visualise CDR3 length distributions
4. Compute and visualise per-position amino acid composition

---

## 1. Reading sequences

`read_fasta()` parses any standard FASTA file into a tibble with one row per
sequence. Multi-line sequences are automatically collapsed.

```{r load}
library(abseqr)

fasta_file <- system.file("extdata", "example_nanobodies.fasta",
                           package = "abseqr")
seqs <- read_fasta(fasta_file)
seqs
```

The `length` column gives the total sequence length in residues. Our example
set contains `r nrow(seqs)` nanobody sequences.

---

## 2. Extracting CDR3 regions

`extract_cdr3()` uses conserved flanking residues to identify CDR3 boundaries.
For VHH nanobodies, CDR3 is flanked by a conserved **cysteine (C)** at the
N-terminal end and a **WGXG** motif at the C-terminal end — following the
Kabat/IMGT convention used in structural databases such as SAbDab.

```{r cdr3}
seqs <- extract_cdr3(seqs)
seqs |> dplyr::select(header, cdr3, cdr3_length)
```

Sequences where no CDR3 could be detected receive `NA`. We can check how
many were successfully extracted:

```{r cdr3-summary}
n_total <- nrow(seqs)
n_found <- sum(!is.na(seqs$cdr3))
cat(sprintf("CDR3 extracted: %d / %d sequences (%.0f%%)\n",
            n_found, n_total, 100 * n_found / n_total))
```

---

## 3. CDR3 length distribution

CDR3 length is a key characteristic of nanobodies. Nanobody CDR3s tend to be
longer than their conventional antibody counterparts (typically 16–24 residues
in VHHs vs 10–15 in VH domains), partly because they often form protruding
loop structures that reach into enzyme active sites.

```{r length-dist, fig.cap="CDR3 length distribution across example nanobodies."}
cdr3_length_dist(seqs,
                 subtitle = paste0("n = ", n_found, " nanobodies"))
```

---

## 4. Per-position amino acid composition

To analyse position-specific amino acid preferences, we first select sequences
of a single CDR3 length (the most common length in the dataset), then compute
composition with `aa_composition()`.

```{r most-common-length}
# Find the most common CDR3 length
common_len <- seqs |>
  dplyr::filter(!is.na(cdr3_length)) |>
  dplyr::count(cdr3_length, sort = TRUE) |>
  dplyr::slice(1) |>
  dplyr::pull(cdr3_length)

cat("Most common CDR3 length:", common_len, "residues\n")
```

```{r composition}
comp <- aa_composition(seqs, length_filter = common_len)
comp
```

`aa_composition_plot()` visualises this as a heatmap where darker cells
indicate higher frequency at that position.

```{r composition-plot, fig.height=6, fig.cap="Per-position amino acid composition for the most common CDR3 length. Amino acids are grouped by physicochemical property."}
aa_composition_plot(
  comp,
  title = paste0("CDR3 Amino Acid Composition (length = ", common_len, ")")
)
```

Position-specific patterns reflect selection pressures on antigen binding:
positions near the tips of the CDR3 loop tend to show more diversity, while
positions flanking the conserved cysteines are more constrained.

---

## Summary

| Function | Input | Output |
|---|---|---|
| `read_fasta()` | FASTA file path | Tibble: header, sequence, length |
| `extract_cdr3()` | Tibble or character vector | Adds cdr3, cdr3_length columns |
| `cdr3_length_dist()` | Tibble with cdr3_length | ggplot bar chart |
| `aa_composition()` | Tibble with cdr3 | Tibble: position × amino_acid frequencies |
| `aa_composition_plot()` | aa_composition tibble | ggplot heatmap |

For further analysis, the tidy output of each function is designed to plug
directly into `dplyr` and `ggplot2` workflows.

---

## Session info

```{r session}
sessionInfo()
```
